---
title: "Lab #20 - Panel Data (MLDA Revisited)"
author: "Econ 224"
date: "November 15th, 2018"
---

<!-- knitr global options -->
```{r, include = FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=4.5, fig.height=3.5, fig.align = 'center')
```

# Introduction 
This lab revisits the MLDA example using panel data methods (state and year effects) rather than regression discontinuity.
Before beginning, please download the file `deaths.dta` from the Mastering 'Metrics website under "Killer Apps > Chapter 5."
Here is a description of the variables that you will need for this exercise:


| Name | Description |
|---------------|-----------------------------------------------|
| `state` | Indicator for US States and DC |
| `year` | Year |
| `pop` | Population in state $s$ in year $t$ |
| `legal` | Prop. of 18-20 year olds who can legally drink in state $s$ in year $t$ |
| `agegr` | Indicator for age ranges (2 = 18-20 year olds) |
| `mrate` | Mortality rate in state $s$ in year $t$ |
| `dtype` | Indicator for *which* mortality category `mrate` contains (1 = all deaths) |
| `beertaxa` | Measure of per-unit beer taxes in state $s$ in year $t$ |

# Exercises

1. Preliminaries:
      (a) Use an appropriate package to open `deaths.dta` in R.
      (b) Convert `year` to factor using `as.factor` 
      (c) Use `as.factor` to create a new variable called `year_factor` containing the same information as `year` but stored as a factor.  
      (c) Restrict the sample to years before 1984, 18-20 year olds, and "all deaths" mortality rates.
2.   
      (a) Use `lm_robust` to estimate the effect of `legal` on `mrate` including state and year effects. Use cluster robust standard errors by setting `clusters = state` and `se_type = 'stata'`. 
      (b) Repeat (a), but run a *weighted* regression by setting `weights = pop`.
      (c) Repeat (b) but allow for *state-specific* effects by including an interaction between `state` and `year`. Why is this different from including an interaction between `state` and `year_factor`?
      (d) Come up with an appropriate way to display *only* the coefficient estimates and standard errors for `legal`, and not all the estimates of state and year effects. Discuss your findings.

3. Repeat 2, but control for beer taxes. Discuss your findings.


# Solutions
<!-- ANS_START -->
```{r, message = FALSE}

# 1- Preliminaries
library(tidyverse)
library(haven)
library(estimatr)

mlda <- read_dta('~/econ224/labs/deaths.dta')
mlda <- mlda %>%
  filter(year <= 1983, agegr == 2, dtype == 1) %>%
  mutate(year_factor = factor(year), state = factor(state))

# 2
reg1 <- lm_robust(mrate ~ legal + state + year_factor - 1, 
                  data = mlda, clusters = state, se_type = 'stata')
reg2 <- lm_robust(mrate ~ legal + state + year_factor + state:year - 1, 
                  data = mlda, clusters = state, se_type = 'stata')
reg3 <- lm_robust(mrate ~ legal + state + year - 1, 
                  data = mlda, weights =  pop, clusters = state, se_type = 'stata')

# 3
reg4 <- lm_robust(mrate ~ legal + beertaxa + state + year_factor - 1, 
                  data = mlda, clusters = state, se_type = 'stata')
reg5 <- lm_robust(mrate ~ legal + beertaxa + state + year_factor + state:year - 1, 
                  data = mlda, clusters = state, se_type = 'stata')
reg6 <- lm_robust(mrate ~ legal + beertaxa + state + year - 1, 
                  data = mlda, weights =  pop, clusters = state, se_type = 'stata')

# Results
estimates <- c(coef(reg1)[1], coef(reg2)[1], coef(reg3)[1],
               coef(reg4)[1], coef(reg5)[1], coef(reg6)[1])
std_errors <- c(reg1$std.error[1], reg2$std.error[1], reg3$std.error[1],
                reg4$std.error[1], reg5$std.error[1], reg6$std.error[1])

results <- cbind(estimates, std_errors)
row.names(results) <- paste0('reg', 1:6)
results
```
<!-- ANS_END -->
