---
title: "Lab #5 - Predictive Regression I"
author: "Econ 224"
date: "September 11th, 2018"
---

<!-- knitr global options -->
```{r, include = FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=4.5, fig.height=3.5, fig.align = 'center')
```

# Introduction 
This lab provides a crash course on least squares regression in R.
In the interest of time we'll work with a very simple, but somewhat boring, dataset that requires very little explanation. 
In our next lab and on the problem set you'll use what you've learned here to look at much more interesting examples!

# The `mtcars` Dataset
The built-in R dataset `mtcars` contains information on 32 models of automobile from 1973-74 as reported in *Motor Trend Magazine*.
For more information on the variables, see the R help file `?mtcars`.
Note that `mtcars` is a dataframe rather than a tibble.
Just to keep things simple I won't convert it to a tibble.
But don't worry: everything I demonstrate in this tutorial will work just as well with a tibble as with a dataframe.
A tibble *is* a dataframe even though a dataframe is not a tibble.
(C.f. a square is a rectangle, but a rectangle is not a square.)
Here are the first few rows of the `mtcars`: 
```{r}
head(mtcars)
```
Our goal will be to predict `mpg` (fuel efficiency in miles/gallon) using the other variables such as `cyl` (\# of cylinders), `disp` (engine displacement in cubic inches), `hp` (horsepower), and `wt` (weight in thousands of pounds).

# The `lm` Command
The command for least squares regression in R is `lm` which stands for *linear model*.
The basic syntax is as follows: `lm([Y variable] ~ [1st predictor] + ... + [pth predictor], [dataframe])`.
For example, to predict `mpg` using `disp` and `hp` we would run the command
```{r}
lm(mpg ~ disp, mtcars)
```

# Exercise \#1
Carry out a regression predicting `mpg` using `disp`, `hp`, `cyl` and `wt`
*Write your code and solutions here.*


## Getting More Information from `lm`
If we simply run `lm` as above, R will display only the estimated regression coefficients: $\widehat{\beta}_0, \widehat{\beta}_1, \hdots, \widehat{\beta}_p$ along with the command used to run the regression: `Call`.
To get more information, we need to *store* the results of our regression.
```{r}
reg1 <- lm(mpg ~ disp + hp, mtcars)
```
If you run the preceding line of code in the R console, it won't produce any output.
But if you check your R environment after running it, you'll see a new `List` object: `reg1`.
To see what's inside this list, we can use the command `str`:
```{r}
str(reg1)
```
Don't panic: you don't need to know what all of these list elements are.
The important thing to understand is that `lm` returns a *list* from which we can extract important information about the regression we have run.
To extract the regression coefficient estimates, we use `coef`
```{r}
coef(reg1)
```
To extract the regression residuals, we use `resid`
```{r}
resid(reg1)
```
and to extract the *fitted values* i.e. the predicted values of $Y$, we use `fitted.values`
```{r}
fitted.values(reg1)
```

# Exercise \# 2
1. Plot a histogram of the residuals from `reg1` using `ggplot` with a bin width of 1.25. Is there anything noteworthy about this plot?
2. Calculate the residuals "by hand" by subtracting the fitted values from `reg1` from the column `mpg` in mtcars. Use the R function `all.equal` to check that this gives the same result as `resid`.

# Solution to Exercise \#2
*Write your code and solutions here.*

# Summarizing Regression Output
To view the usual summary of regression output, we use the `summary` command:
```{r}
summary(reg1)
```
Among other things, `summary` shows us the coefficient estimates and associated standard errors.
It also displays the t-value (Estimate / SE) and associated p-value for a test of the null hypothesis $H_0\colon \beta = 0$ versus $H_1\colon \beta \neq 0$.
Farther down in the output, `summary` provides the residual standard error and R-squared.
It turns out the summary command *itself* returns a list.
In particular,
```{r}
str(summary(reg1))
```
This fact can come in handy when you want to *extract* some of the values from the regression summary table to use for some other purpose.
For example, we can display *only* the R-squared as follows:
We could do this as follows:
```{r}
summary(reg1)$r.squared
```
and only the F-statistic with its associated degrees of freedom as follows:
```{r}
summary(reg1)$fstatistic
```

# Exercise \#3
1. Use `summary` to display the results of the regression you ran in Exercise \#1 above.
2. Figure out how to extract and display *only* the regression standard error from the results of `summary` in part 1 of this exercise.
3. Calculate the regression standard error for the regression from part 1 of this exercise "by hand" and make sure that your answer matches part 2. Hint: use `resid`

# Solution to Exercise \#3
*Write your code and solutions here.*

# Regression Without an Intercept
More than 99% of the time, it makes sense for us to include an intercept $\beta_0$ in a linear regression.
To see why, consider the meaning of $\beta_0$: this is the value of $Y$ that we would predict if $X_1 = X_2 = \hdots = X_p = 0$.
Unless we have some very strong *a priori* knowledge, there is no reason to suppose that the mean of $Y$ should be zero when all of the predictors are zero.
In some very special cases, however, we *do* have such special knowledge.
To *force* the intercept in a regression to be zero we use the syntax `-1`, for example
```{r}
summary(lm(mpg ~ disp - 1, mtcars))
```

# Exercise \#4
What do you get if you run the regression `lm(mpg ~ 1, mtcars)`?

# Solution to Exercise \#4
*Write your code and solutions here.*


# F-tests
Suppose we want to test the *joint* null hypothesis $H_0\colon \beta_1 = \beta_2 = \cdots = \beta_p = 0$ versus the alternative that at least one of these coefficients is non-zero.
This is equivalent to testing the null hypothesis that none of the predictors $X_1, \cdots, X_p$ is helpful in predicting $Y$.
This test is automatically carried out by `summary`.
Consider a regression that uses `disp`, `hp`, `wt` and `cyl` to predict `mpg`
```{r}
reg2 <- lm(mpg ~ disp + hp + wt + cyl, mtcars)
summary(reg2)
```
At the very bottom of the `summary` output is a line that begins `F-statistic`.
This line contains the results of the F-test of the joint null described above.
In this case the p-value is miniscule: there is very strong evidence that at least one of the predictors is helpful in predicting `mpg`.
To get a better understanding of what's involved here, we can calculate the p-value "by hand" as follows:
```{r}
summary(reg2)$fstatistic
curve(expr = df(x, 4, 27), 0, 40, n = 1001,
      ylab = 'density',
      main = 'Is 37.8 an F(2,29) random draw?')
abline(v = 37.84413, lty = 2, col = 'red')
1 - pf(37.84413, 4, 27)
```
Under the null hypothesis, the F-statistic is a draw from an $F$ random variable with numerator degrees of freedom $2$ and denominator degrees of freedom $29$.
(If you are unfamiliar with the F-distribution see my Tutorial [Friends of the Normal Distribution](http://ditraglia.com/Econ103Public/Rtutorials/friends_of_normal.html).)
So is is plausible that the value 37.8 came from an $F(2,29)$ distribution?
From my plot of the corresponding density function, the answer is clearly *no*.
We have very strong evidence against the null hypothesis.

Sometimes we only want to test the null that a *subset* of our predictors is unhelpful for predicting $Y$.
For example, in `reg2` we might ask whether `wt` and `cyl` provide *extra* information for predicting `mpg` after we have already included `disp` and `hp` in our model.
To carry out this test, we use the function `linearHypothesis` from the package `car`.
Make sure to install this package before proceeding.
Note the syntax: `linearHypothesis([lm object], c('[first restriction]', ..., '[last restriction]')`
```{r}
library(car)
linearHypothesis(reg2, c('wt = 0', 'cyl = 0'))
```
The two key numbers to look for in the output are `F`, the value of the F-statistic, and `Pr(>F)`, the p-value.
The other values are the inputs used to calculate `F`.
(See Equation 3.24 in ISL.)
In this instance we strongly reject the null hypothesis that `wt` and `cyl` are irrelevant for predicting `mpg` after controlling for `disp` and `hp`.

# Exercise \#5
Generate two vectors of independent standard normal draws `x` and `z`.
Each vector should contain as many elements as there are rows in `mtcars`.
Use the command `set.seed(1234)` before making your random draws so that they are replicable.
(By first setting the seed to a fixed number, you ensure that the same random draws will be made any time that you re-run this code chunk.) 
Carry out a new regression, `reg3`, that *augments* `reg2` by adding the predictors `x` and `z`.
Then carry out an F-test the null hypothesis that `x` and `z` are irrelevant for predicting `mpg` after controlling for `disp`, `hp`, `wt`, and `cyl`.
Interpret your findings.
Do the results of the test make sense?

# Solution to Exercise \#5
*Write your code and solutions here.*



# Plotting the Regression Line
To get an idea of whether our regression model looks reasonable, it's always a good idea to make some plots.
When we have a single predictor $X$, it is common to plot the raw $X$ and $Y$ observations along with the regression line.
It's easy to do this using `ggplot`.
Suppose we wanted to predict `mpg` using `disp`.
Here's the `ggplot` way to plot the data and regression line:
```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() +
  geom_smooth(method = 'lm')
```
Notice that I specified `aes` inside of `ggplot`.
This ensures that both `geom_point` and `geom_smooth` "know" which variable is `x` and which variable is `y`.
Notice moreover, that the `ggplot` way of doing this includes *error bounds* for the regression line.
This is a handy way of visualizing the uncertainty in the line we've fit.


# Exercise \#6
Make a `ggplot` with `hp` on the x-axis and `mpg` on the y-axis that includes the regression line for predicting `mpg` from `hp`.

# Solution to Exercise \#6
*Write your code and solutions here.*

# Polynomial Regression
In your next reading assignment, you'll learn about *polynomial regression*.
The "linear" in linear regression does not actually refer to the relationship between $Y$ and the predictors $X$; it refers to the relationship between $Y$ and the *coefficients* $\beta_0, \beta_1, ..., \beta_p$.
In the expression $Y = \beta_0 + \beta_1 X + \epsilon$, $Y$ is a linear function of $\beta_0$ and $\beta_1$ and it is *also* a linear function of $X$.
In the expression $Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \epsilon$, $Y$ is *still* a linear function of the coefficients, but a *quadratic* function of $X$.
This is a simple example of polynomial regression, which allows us to model more complicated relationships between $X$ and $Y$.
Notice, for example, that the relationship between `mpg` and `disp` looks like it might be curved. 
To accommodate such a relationship, let's try a polynomial regression that includes includes `disp` and `disp^2`.
To do this we use the syntax `I([some transformation of a predictor)`
```{r}
reg3 <- lm(mpg ~ disp + I(disp^2), mtcars)
summary(reg3)
```
Notice that the coefficient on the quadratic term is *highly* statistically significant, which is strong evidence of curvature in the relationship between `mpg` and `disp`.
We can plot the polynomial regression as follows:
```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() +
  geom_smooth(method = 'lm', formula = y ~ x + I(x^2))
```
Notice that this requires us to specify the `formula` argument so that `ggplot` knows that we want to plot a quadratic relationship.

# Exercise \#7
In my code above, I considered a quadratic relationship between `mpg` and `disp`.
Add a cubic term to the regression, plot the points and regression function, and display the results using `summary`.
Comment on the results.

# Solution to Exercise \#7
*Write your code and solutions here.*


# Interaction Effects
An idea closely related to polynomial regression that will also be discussed in your next reading assignment is that of an *interaction*. 
In the model $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_3 + \epsilon$, $Y$ is a linear function of $\beta_0, \beta_1, \beta_2$, and $\beta_3$ but a *nonlinear* function of $X_1$ and $X_2$.
The term $X_1 \times X_2$ is called an *interaction*.
To run a regression with an interaction, we use the syntax `[One Predictor]:[Another Predictor]` for example
```{r}
lm(mpg ~ disp + hp + disp:hp, mtcars)
```

# Exercise \#8 
Fit a regression using `disp`, `disp^2`, `wt`, `wt^2` and the interaction between `wt` and `disp` to predict `mpg` and display the coefficient estimates.


# Solution to Exercise \#8
*Write your code and solutions here.*
 



# Predicting New Observations
To predict new observations based on a fitted linear regression model in R, we use the `predict` function.
For example, consider three hypothetical cars with the following values of displacement and horsepower
```{r}
mydat <- data.frame(disp = c(100, 200, 300),
                    hp = c(150, 100, 200))
mydat
```
Based on the results of `reg1`, we would predict that these cars have the following values of `mpg`
```{r}
predict(reg1, mydat)
```
Note the syntax: the first argument of `predict` is a set of regression results while the second is a dataframe (or tibble) with column names that *match* the variables in the regression we carried out.

# Exercise \#9
1. Check the predictions of `predict` in the preceding chunk "by hand" using `mydat`, `coef`, and `reg1`.
2. Consider three cars with `disp` equal to 125, 175, and 225, respectively.
   Predict `mpg` for each of these based on the regression from Exercise \#7.

# Solution to Exercise \#9
*Write your code and solutions here.*

